---
title: "California ISO Electric Grid Data Analysis"
author: "Michaela Palmer, Melissa Ferriter"
output: github_document
---

```{r setup, include=T}
knitr::opts_chunk$set(echo = T, warning = F, message = F)
```

```{r results='hide'}
libs <- c("tidyverse", "readxl", "ggpubr")
sapply(libs, require, character.only=T)
```

## Read in data

The data used in this analsyis comes from [California Independent System Operator Corporation (CAISO)](http://www.caiso.com/green/renewableswatch.html), which aggregates grid data from electricity producers in California.

```{r}
# other tested regex methods
#sapply(strsplit(filelist, "[/_]"), "[", 6)
#sub(".*(\\d{8}).*", "\\1", filelist)

#sub(".*/", "", sub("_.*", "", filelist))

#sub("_.*", "", basename(filelist))

```

```{r}
jan.data <-
  c(
    "http://content.caiso.com/green/renewrpt/20170115_DailyRenewablesWatch.txt",
    "http://content.caiso.com/green/renewrpt/20170116_DailyRenewablesWatch.txt",
    "http://content.caiso.com/green/renewrpt/20170117_DailyRenewablesWatch.txt",
    "http://content.caiso.com/green/renewrpt/20170118_DailyRenewablesWatch.txt",
    "http://content.caiso.com/green/renewrpt/20170119_DailyRenewablesWatch.txt",
    "http://content.caiso.com/green/renewrpt/20170120_DailyRenewablesWatch.txt",
    "http://content.caiso.com/green/renewrpt/20170121_DailyRenewablesWatch.txt",
    "http://content.caiso.com/green/renewrpt/20170122_DailyRenewablesWatch.txt"
  )
feb.data <-
  c(
    "http://content.caiso.com/green/renewrpt/20170215_DailyRenewablesWatch.txt",
    "http://content.caiso.com/green/renewrpt/20170216_DailyRenewablesWatch.txt",
    "http://content.caiso.com/green/renewrpt/20170217_DailyRenewablesWatch.txt",
    "http://content.caiso.com/green/renewrpt/20170218_DailyRenewablesWatch.txt",
    "http://content.caiso.com/green/renewrpt/20170219_DailyRenewablesWatch.txt",
    "http://content.caiso.com/green/renewrpt/20170220_DailyRenewablesWatch.txt",
    "http://content.caiso.com/green/renewrpt/20170321_DailyRenewablesWatch.txt",
    "http://content.caiso.com/green/renewrpt/20170422_DailyRenewablesWatch.txt"
  )

june.data <-
  c(
    "http://content.caiso.com/green/renewrpt/20170615_DailyRenewablesWatch.txt",
    "http://content.caiso.com/green/renewrpt/20170616_DailyRenewablesWatch.txt",
    "http://content.caiso.com/green/renewrpt/20170617_DailyRenewablesWatch.txt",
    "http://content.caiso.com/green/renewrpt/20170618_DailyRenewablesWatch.txt",
    "http://content.caiso.com/green/renewrpt/20170619_DailyRenewablesWatch.txt",
    "http://content.caiso.com/green/renewrpt/20170620_DailyRenewablesWatch.txt",
    "http://content.caiso.com/green/renewrpt/20170621_DailyRenewablesWatch.txt",
    "http://content.caiso.com/green/renewrpt/20170622_DailyRenewablesWatch.txt"
  )

july.data <-
  c(
    "http://content.caiso.com/green/renewrpt/20170715_DailyRenewablesWatch.txt",
    "http://content.caiso.com/green/renewrpt/20170716_DailyRenewablesWatch.txt",
    "http://content.caiso.com/green/renewrpt/20170717_DailyRenewablesWatch.txt",
    "http://content.caiso.com/green/renewrpt/20170718_DailyRenewablesWatch.txt",
    "http://content.caiso.com/green/renewrpt/20170719_DailyRenewablesWatch.txt",
    "http://content.caiso.com/green/renewrpt/20170720_DailyRenewablesWatch.txt",
    "http://content.caiso.com/green/renewrpt/20170721_DailyRenewablesWatch.txt",
    "http://content.caiso.com/green/renewrpt/20170722_DailyRenewablesWatch.txt"
  )

oct.data <-
  c(
    "http://content.caiso.com/green/renewrpt/20171015_DailyRenewablesWatch.txt",
    "http://content.caiso.com/green/renewrpt/20171016_DailyRenewablesWatch.txt",
    "http://content.caiso.com/green/renewrpt/20171017_DailyRenewablesWatch.txt",
    "http://content.caiso.com/green/renewrpt/20171018_DailyRenewablesWatch.txt",
    "http://content.caiso.com/green/renewrpt/20171019_DailyRenewablesWatch.txt",
    "http://content.caiso.com/green/renewrpt/20171020_DailyRenewablesWatch.txt",
    "http://content.caiso.com/green/renewrpt/20171021_DailyRenewablesWatch.txt",
    "http://content.caiso.com/green/renewrpt/20171022_DailyRenewablesWatch.txt"
  )


```

```{r }
import <- function(data) {
  data.frame(date = as.Date(sub("_.*", "", basename(data)), "%Y%m%d"), read_table2( 
    data,
    col_names = c(
      "hour",
      "geothermal",
      "biomass",
      "biogas",
      "small_hydro",
      "wind",
      "solar_pv",
      "solar_thermal"
    ),
    skip = 2,
    n_max = 24
  )) 
}

jan <- do.call("rbind", lapply(jan.data, import))
feb <- do.call("rbind", lapply(feb.data, import))
june <- do.call("rbind", lapply(june.data, import))
july<- do.call("rbind", lapply(july.data, import))
oct <- do.call("rbind", lapply(oct.data, import))

```

# Daily average data plots

The CAISO data are provided at hourly intervals. Plotting the data for the entire 6-year period would have generated a very messy graph, so I calculated daily means and plotted them instead.


```{r}

```

